---
title: "Project Proposal"
subtitle: "due October 11, 2021 by 11:59 PM "
author: "Harris Upchurch, Biniam Garomsa, Philemon Hailemariam"
date: "10/11/2021"
output: pdf_document
---

# Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse) 
library(data.table)
``` 

# Load Data
```{r load-data, message = F}
oxford_lockdown<-read.csv(file = '~/R/TBD2/data/OxCGRT_US_latest.csv')
flu<-read.csv(file = '~/R/TBD2/data/State_Custom_Data_Influenza.csv')
```

# Introduction and Data, including Research Questions

How have the coronavirus and corresponding prevention measures affected the spread of other infectious diseases, specifically the flu, in the United States? Our first dataset is from the Blavatnik School of Government at Oxford, which collects its information via volunteers monitoring policy announcements. Its variables are different types of COVID prevention measures (banning public gatherings, mask mandates, etc.) and the observations measure the severity of each policy and whether or not they are in place. Each variable is measured for each state and day from January 01, 2020 to today (10/11). This dataset will help us to compare the prevention measures against COVID to levels of infectious diseases from other datasets, which will allow us to see if these preventions affected other infectious diseases. This could help us to understand how to allocate resources in the future, and if we should expect infectious disease deaths to be reduced due to effective measures, which would mean that estimating COVID from excess deaths based on previous years would be an underestimation. Our second group of datasets is from the CDC and includes information on the deaths in each week and state for the flu, COVID, and pneumonia. This will allow us to convert each flu death number to a death rate for its state and compare it to stringency for each date. These data were collected by the CDC by surveying hospitals across the country each year. Additionally, these data for the 2021-2022 season are provisional and may change.

# Glimpse

```{r glimpse}
glimpse(oxford_lockdown)
glimpse(flu)
```


# Data Analysis Plan

The outcome variable is the deaths due to non-COVID-19 infectious diseases over the given time period, which will be compared based on the different policies of different states. The predictor will be the stringency index of prevention measures of different states, a variable generated from combination of different policies by the Oxford dataset. Statistical methods that may prove useful in answering our question are generating a ratio of mean or total stringency to ID death rates for each state and finding a correlation, creating line graphs comparing stringency and deaths for several states, or creating a graph showing lines for each state over time of the ratio between deaths and stringency for each day or week. Data that would support our hypothesis that prevention measures suppress other infectious diseases as well would show an correlation between higher stringency and lower deaths, which could be found by comparing different states and different times. Currently, we are comparing stringency to flu deaths for each week and state where data is available. For the data to be readable, the deaths in each state will have to be converted to a rate as a proportion of the population, and states or times will likely have to be faceted to reduce the number of data points on a single graph. Alpha or color could potentially be used to show time differences.
```{r initialvis}
flu<-flu%>%
  transform(RegionName = SUB.AREA)%>%
  mutate(year = ifelse('season' == 2021-2022, 2021, ifelse('season' == 2020-2021, 2020, 2019)))%>%
  transform(yearweek = as.numeric(paste(year, 'week', sep = "")))
oxford_lockdown<-oxford_lockdown%>%
  mutate(year = year(as.numeric('Date')), week = week(as.numeric('Date')))%>%
  mutate(yearweek = as.numeric(paste('year', 'week', sep = "")))
combdat<-left_join(oxford_lockdown, flu, by = 'RegionName','yearweek')
```
``` {r ggplot}
combdat%>%
  filter('date' >= 20210901)%>%
  ggplot(aes(x = StringencyIndex, y = NUM.INFLUENZA.DEATHS)) +
  geom_point()
#collapse by week, convert state deaths to rate/proportional to population
#yearweek not working
```

